---
title: "Data wrangling I"
format: 
  html: default
code-annotations: select
---

```{r}
library(tidyverse)

nhanes <- read.csv("https://raw.githubusercontent.com/ccb-hms/hsdm-r-course/main/session-materials/session0/session0Data.csv")
```

## Mutate

Frequently you'll want to create new columns based on the values of existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

To create a new column of age in months:

```{r, purl=TRUE}
nhanes %>%
  mutate(age.months = age.years * 12) %>%
  select(age.months, age.years) %>%
  head()
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r, purl=TRUE}
nhanes %>%
  mutate(age.months = age.years * 12,
         lived_200_months = age.months >= 200) %>%
  select(age.months, age.years, lived_200_months)
```

::: {.callout-note icon=false}

## Challenge

Create a new data frame from the `nhanes` data that meets the following
criteria: contains the columns `sex`, `under_14`,
`ethnicity`, and `family.PIR` columns. The `under_14` column should be a logical variable indicating whether or not the age of the participant is less than 14. This data frame must
only contain participants born in the USA or Mexico, have a non-missing plasma glucose value, and with a BMI less than 30.

:::

::: {.callout-tip icon=false collapse=true}

## Solution

```{r, eval=TRUE, purl=TRUE}
nhanes %>%
  mutate(under_14 = age.years < 14) %>%
  filter(birthplace == 'Born in 50 US States or Washi' | birthplace == "Born in Mexico") %>%
  filter(!is.na(plasma.glucose)) %>%
  filter(bmi < 30) %>%
  select(sex, under_14, ethnicity, family.PIR)
```

:::

# Data

This week we are going to start learning how to get to the cleaned data we have been using for the past two weeks. 

Note that this week the data is in the `.RData` format.
The `.RData` format is specific to R and allows us to easily dump multiple objects to a file exactly as the are inside of our R session. 
This format is not recommended for research as it limits analyses to only using R and could potentially not be compatible with a future R version. 
However, for today it will allow us to jump right into the data cleaning process. 

The `nhanes_beheshti_data.RData` file contains 3 separate dataframes, `base_df_d`, `base_df_e`, and `base_df_f`.
They each contain the equivalent set of columns from the `D`, `E`, and `F` NHANES data files for 2005-2006, 2007-2008, and 2009-2010, respectively.  

Let's load and take a look at the data.

```{r}
load("../data/nhanes_beheshti_data.RData")
```
```{r}
summary(base_df_d)
```
```{r}
summary(base_df_e)
```
```{r}
summary(base_df_f)
```

## Split-apply-combine data analysis

Many data analysis tasks can be approached using the
*split-apply-combine* paradigm: split the data into groups, apply some
analysis to each group, and then combine the results. **`dplyr`**
makes this very easy through the use of the `group_by()` function.

```{r}
nhanes %>%
  group_by(birthplace)
```

The `group_by()` function doesn't perform any data processing, it
groups the data into subsets: in the example above, our initial
`tibble` of `r nrow(nhanes)` observations is split into
`r length(unique(nhanes$birthplace))` groups based on the `birthplace` variable.

We could similarly decide to group the tibble by sex:

```{r}
nhanes %>%
  group_by(sex)
```

Here our initial `tibble` of `r nrow(nhanes)` observations is split into
`r length(unique(nhanes$sex))` groups based on the `sex` variable.

Once the data has been grouped, subsequent operations will be
applied on each group independently.

### The `summarise()` function

`group_by()` is often used together with `summarise()`, which
collapses each group into a single-row summary of that group.

`group_by()` takes as arguments the column names that contain the
**categorical** variables for which you want to calculate the summary
statistics. So to compute the mean `bmi` by birthplace:

```{r}
nhanes %>%
  group_by(birthplace) %>%
  summarise(mean_bmi = mean(bmi, na.rm = TRUE))
```

But we can can also group by multiple columns:

```{r}
nhanes %>%
  group_by(sex, ethnicity) %>%
  summarise(mean_bmi = mean(bmi, na.rm = TRUE))
```

Once the data is grouped, you can also summarise multiple variables at the same
time (and not necessarily on the same variable). For instance, we could add a
column indicating the median `plasma.glucose` by sex and ethnicity:

```{r, purl=TRUE}
nhanes %>%
  group_by(sex, ethnicity) %>%
  summarise(mean_plasma_glucose = mean(plasma.glucose),
            median_plasma_glucose = median(plasma.glucose))
```

::: {.callout-note icon=false}

## Challenge

Calculate the mean `hba1c` of participants born in the USA by age.

:::


::: {.callout-tip icon=false collapse=true}

## Solution

```{r, purl=TRUE}
nhanes %>%
  filter(birthplace == 'Born in 50 US States or Washi') %>%
  group_by(age.years) %>%
  summarise(mean_hba1c = mean(hba1c))
```

:::

### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each age, we would do:

```{r, purl=TRUE}
nhanes %>%
    count(age.years)
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarising it by counting the number of observations in that group. In other words, `nhanes %>% count(age.years)` is equivalent to:

```{r, purl=TRUE}
nhanes %>%
    group_by(age.years) %>%
    summarise(n = n())
```

The previous example shows the use of `count()` to count the number of rows/observations
for *one* factor (i.e., `infection`).
If we wanted to count a *combination of factors*, such as `age` and `sex`,
we would specify the first and the second factor as the arguments of `count()`:

```{r, purl=TRUE}
nhanes %>%
    count(age.years, sex)
```

It is sometimes useful to sort the result to facilitate the comparisons.
We can use `arrange()` to sort the table.
For instance, we might want to arrange the table above by age:

```{r, purl=TRUE}
nhanes %>%
    count(age.years, sex) %>%
    arrange(age.years)
```

or by counts:

```{r, purl=TRUE}
nhanes %>%
    count(age.years, sex) %>%
    arrange(n)
```

To sort in descending order, we need to add the `desc()` function:

```{r, purl=TRUE}
nhanes %>%
    count(age.years, sex) %>%
    arrange(desc(n))
```

::: {.callout-note icon=false}

## Challenge

1. How many participants have a non-empty plasma glucose value for each age?
2. Use `group_by()` and `summarise()` to evaluate the poverty income ratio (`family.PIR`) by ethnicity. Which ethnicity has the highest poverty income ratio?

:::

::: {.callout-tip icon=false collapse=true}

## Solution

```{r}
#1
nhanes %>% filter(!is.na(plasma.glucose)) %>%
  group_by(age.years) %>%
  count()

#2
nhanes %>% group_by(ethnicity) %>%
  summarise(mean_PIR = mean(family.PIR, na.rm = TRUE)) %>%
  arrange(desc(mean_PIR))
```

:::

# Combining data

In the reading, we saw how to combine multiple sets of observations by appending datasets. However, before combining multiple datasets we need to make sure corresponding columns between years can be combined. 

NHANES will often change how it encodes or records different variables from year to year.
Our first red flag are any columns whose name has changed.
Let's take a look at the column names.

```{r}
colnames(base_df_d)
colnames(base_df_e)
colnames(base_df_f)
```

We can also check for equality directly:

```{r}
colnames(base_df_d)[colnames(base_df_d) != colnames(base_df_e)]
colnames(base_df_e)[colnames(base_df_e) != colnames(base_df_f)]
colnames(base_df_f)[colnames(base_df_f) != colnames(base_df_d)]
```

It looks like `DMDBORN` changed after 2006 to `DMDBORN2`. 
If we check the demographic variable definitions for the three collection cycles [DEMO_D](https://wwwn.cdc.gov/nchs/nhanes/2005-2006/DEMO_D.htm#DMDBORN), [DEMO_E](https://wwwn.cdc.gov/Nchs/Nhanes/2007-2008/DEMO_E.htm#DMDBORN2), and [DEMO_F](https://wwwn.cdc.gov/Nchs/Nhanes/2009-2010/DEMO_F.htm#DMDBORN2) we can see that the encoding has changed over time.

To continue the analysis, let's also convert these dataframes into tibbles so we can take advantage of Tidyverse functionality. 

```{r}
tib_d <- as_tibble(base_df_d)
tib_e <- as_tibble(base_df_e)
tib_f <- as_tibble(base_df_f)
```

Finally, let's also view the metadata of the columns.

```{r}
DT::datatable(metadata)
```


::: {.callout-tip appearance="minimal"}
## Exercise
a. Resolve any changes in variable coding between the datasets.
Use your own judgement on the best way to resolve any issues (doing nothing might also be a solution). 

```{r}

# Step 1: If needed, change things to have the same sets of values
# TODO your code here

# Step 2: Rename the DMDBORN2 columns to match between the datasets
tib_e <- rename(tib_e, DMDBORN = DMDBORN2)
tib_f <- rename(tib_f, DMDBORN = DMDBORN2)
```

b. Check table 1b to see how the authors chose to handle this in [Beheshti et. al. 2021](https://pubmed.ncbi.nlm.nih.gov/33892837/). Do you think there were any other factors which went into their decision?

:::

## Combining the data

Now we can combine the data together.

```{r}
# Combine all years together
tib_all <- bind_rows(tib_d, tib_e, tib_f)
```


# Data cleaning pipeline

We can choose to leave the variables with their NHANES column names or change them to be a bit more human-readable. 
Either way, we also want to convert each variable to it's appropriate type or category. 

## Numeric variables
Let's start by converting numeric columns:

```{r}
# Set numeric columns to be numeric
tib_all <- mutate(tib_all,
                   RIDAGEYR = as.numeric(RIDAGEYR),
                   LBXGLU = as.numeric(LBXGLU),
                   LBXGH = as.numeric(LBXGH),
                   BMXBMI = as.numeric(BMXBMI))
```

## Categorical variables

Many of the NHANES values are categorical data, but right now are stored as text. 
We can check what values exist by converting them to factors before making the decision of how to handle them in the analysis: 

```{r}
tib_all |>
  mutate(across(c(RIAGENDR, RIDRETH1, DMDBORN), as.factor), .keep = "none") |>
  summary()
```

It looks like one of the options for `DMDBORN` has a lingering quote character.
This can happen due to irregularities in how NHANES data is presented or small mistakes in data processing. 
We use the `gsub` function to replace all instances of double quotes with the empty string in the column.

```{r}
tib_all <- mutate(tib_all,
                   RIAGENDR = as.factor(RIAGENDR),
                   RIDRETH1 = as.factor(RIDRETH1),
                   DMDBORN = gsub("\"", "", DMDBORN), # Remove quotes
                   DMDBORN = as.factor(DMDBORN),
                   OHXDECAY = (OHXDECAY == "Yes"),
                   OHXREST = (OHXREST == "Yes"))

```

## New variables

We need to create the variables mentioned in the paper are not directly present in NHANES.
For some variables such as country of birth, we could store the categories in `DMDBorn` but this way we can also keep the original distribution.  

```{r}
# Set columns to categories as in the paper
tib_all <- mutate(tib_all,

                    age.cat = cut(
                      RIDAGEYR,
                      breaks = c(13, 15, 18, 100),
                      include.lowest = TRUE,
                      labels = c("13-15", "16-18", "19+")),
                   
                    plasma.glucose.cat = case_when(
                     LBXGLU < 100 ~ "<100 mg/dl",
                     LBXGLU < 126 ~ ">=100 mg/dl and <126 mg/dl", 
                     LBXGLU >= 126 ~ ">=126 mg/dl",
                     .default = NA),
                   
                   hba1c.cat = case_when(
                     LBXGH < 5.7 ~ "<5.7%",
                     LBXGH >= 5.7 ~ ">=5.7% and <6.5%",
                     LBXGH >= 6.5 ~ ">= 6.5%",
                     .default = NA),
                   
                   bmi.cat = case_when( # If we were doing this from scratch we might also want an underweight category
                     BMXBMI < 25 ~ "Normal", 
                     BMXBMI < 30 ~ "Overweight",
                     BMXBMI >= 30 ~ "Obese",
                     .default = NA), 
                   
                   family.PIR.cat = case_when(
                     INDFMPIR == "PIR value greater than or equa" ~ ">= 1",
                     INDFMPIR == "Value greater than or equal to" ~ ">= 1",
                     as.numeric(INDFMPIR) >= 1 ~ ">=1",
                     as.numeric(INDFMPIR) < 1 ~ "<1",
                     .default = NA),
                   
                   birthplace = case_when(
                     DMDBORN == "Born in 50 US States or Washi" ~ "Within the US",
                     is.na(DMDBORN) ~ NA,
                     DMDBORN == "Don't Know" ~ NA,
                     DMDBORN == "Refused" ~ NA,
                    .default = "Outside the US"),
                   dental.caries = OHXDECAY | OHXREST)

```

::: {.callout-tip appearance="minimal"}
## Exercise

Add a `diabetes` column to the dataset based on the guidelines from Beheshti et. al:

>Individuals with an HbA1C of at least 6.5 percent
or a plasma-fasting glucose value of at least 126 mg/dl were
considered diabetics; prediabetics were those whose HbA1C
ranged from 5.7 percent to 6.4 percent and whose fasting
plasma glucose remained within 100 to 125 mg/dl; the remaining
study participants, who had less than 5.7 percent HbA1C
and less than 100 mg/dl fasting plasma glucose, were classified
as nondiabetics.

It is recommended to use `case_when`, but feel free to use any method you want. 

*Hint: remember to account for missing data! When should `diabetes` be `NA`?

```{r}
# Add diabetes column
tib_all <- tib_all |> 
          mutate(diabetes = case_when(
            is.na(LBXGH) & is.na(LBXGLU) ~ NA,
           LBXGH >= 6.5 | LBXGLU >= 126 ~ "diabetic",
           LBXGH >= 5.7 | LBXGLU >= 100 ~ "prediabetic",
            .default = "nondiabetic"))
```
:::



# Adding other diabetes data

While this is the data looked at in the paper, let's take a look at some other diabetes data from the same NHANES cycles. 

```{r}
load("../data/nhanes_diabetes.RData")
```

This contains 4 variables from the DIQ table:

```{r}
DT::datatable(metadata_diabetes)
```

As before, we need to combine the three data cycles together.

```{r}
# This time let's combine the dataframes first
diabetes_all <- rbind(diabetes_d, diabetes_e, diabetes_f)

# And then convert to a tibble
diabetes_all <- as_tibble(diabetes_all) |>
  mutate(DIQ010 = as.factor(DIQ010),
         DIQ160 = as.factor(DIQ160),
         DIQ050 = as.factor(DIQ050)) |>
  select(-c(Begin.Year, EndYear))
```

We now need to combine the diabetes data with the rest of the nhanes data, adding in the new columns. 
This is a more complicated process than simply adding new rows to a dataset, as we need to determine which rows in the diabetes data are referencing the same participant as rows in the main data. 
This process is called joining or merging. 
To merge two datasets, we need to decide what to do with columns which appear in one dataset but not the other. 

We also need to determine which variable in the dataset we want to use as the **key** when joining, the column(s) that identifies which rows in the each dataset we want to match. 
In the `dplyr` join functions this is referred to as the `by` argument. 

In NHANES, we have the `SEQN` variable which uniquely identifies each participant in each survey cycle. 
We can use this as the key. 
In this case we want to perform join where we maintain all rows from the main dataset, but drop any rows which only exist in the diabetes dataset. 
Therefore, let's do a left join (with the main dataset on the left).

```{r}
tib_all <- left_join(tib_all, diabetes_all, by = "SEQN")
summary(tib_all)
```

Let's also check the number of `NA`'s in the new columns to make sure the data has matched up. 

```{r}
#Base R way
na_counts <- colSums(is.na(tib_all))
na_counts <- data.frame(na_counts[order(na_counts, decreasing = TRUE)])
colnames(na_counts) <- c("NA Count")
DT::datatable(na_counts)
```

Unfortunately, it looks `DID040` won't be very useful. 

## Exploring Other Diabetes Variables

We'll start by checking the consistency of the DIQ variables. 
Was there anyone who said they were both diabetic and prediabetic?
Here we're using the `table` function, which will give us a nice counts table as output. 
We'll see some other ways to make tables next week. 

```{r}
table(tib_all[,c("DIQ010", "DIQ160")], exclude = NULL)
```

Now let's see how much the diabetes definition by Beheshti agrees with the `DIQ` variables.

We'll make a variable with the same categories from the `DIQ` data. 
We'll leave out the borderline, don't know, and refused responses. 

```{r}
tib_all <- tib_all |> 
  mutate(diabetes.DIQ = case_when(
    DIQ010 == "Yes" ~ "diabetic",
    DIQ160 == "Yes" ~ "prediabetic",
    DIQ160 == "No" | DIQ010 == "No" ~ "nondiabetic"
  ))

table(tib_all[,c("diabetes.DIQ", "diabetes")])
```