---
title: "Data wrangling I"
format: 
  html: default
code-annotations: select
---

# Synthetic CMS Data

Today we're going to start with the [CMS 2008-2010 Data Entrepreneurs’ Synthetic Public Use File (DE-SynPUF)](https://www.cms.gov/data-research/statistics-trends-and-reports/medicare-claims-synthetic-public-use-files/cms-2008-2010-data-entrepreneurs-synthetic-public-use-file-de-synpuf) to look at combining data. We'll return to this data in much more detail during our next session. 

## Load Beneficiaries summary file

Let's load the data and the column labels.

```{r}
library(tidyverse)
library(DT)

ben_summary_2008 <- read_csv("../data/de_synpuf/DE1_0_2008_Beneficiary_Summary_File_Sample_1.csv")
ben_summary_2009 <- read_csv("../data/de_synpuf/DE1_0_2009_Beneficiary_Summary_File_Sample_1.csv")
ben_summary_labels <- read_csv("../data/de_synpuf/ben_metadata.csv")
summary(ben_summary_2008)
DT::datatable(ben_summary_labels)
```

## Load inpatient table

Now let's load the inpatient data. 

```{r}
inp_data <- read_csv("../data/de_synpuf/DE1_0_2008_to_2010_Inpatient_Claims_Sample_1.csv")
inp_labels <- read_csv("../data/de_synpuf/inp_metadata.csv")
summary(inp_data)
DT::datatable(inp_labels)
```

# Combining Data

There is an overview of joining in the [data transformation cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf).

## Adding more rows


```{r}
ben_summary <- ben_summary_2008[,1:8]
inp_data <- inp_data[,1:20]
```

## Merging datasets by matching values


# Creating new columns

## Data

```{r}
nhanes <- read.csv("../data/nhanes_diabetes.csv")
```



# Data cleaning pipeline

Let's load the raw version of the nhanes data we looked at during the first two sessions. 
This data is stored in the SAS transport `.xpt` format, so we'll read it in using the `read_xpt` function from the `haven` package. `haven` also has a `read_sas` function for reading `sas7bdat` data, as well as functions for reading and writing Stata and SPSS formats. 

```{r}
library(haven)
nhanes_raw <- read_xpt("../data/raw_nhanes.xpt")

```


# String Manipulation


## Combining and joining columns

Let's make a small dataframe to demonstrate splitting and combining on columns. 

```{r}
# Make a small example dataframe

df <- data.frame(
  case_ID = c(1:6),
  symptoms  = c("jaundice, fever, chills",     # patient 1
                "chills, aches, pains",        # patient 2 
                "fever",                       # patient 3
                "vomiting, diarrhoea",         # patient 4
                "bleeding from gums, fever",   # patient 5
                "rapid pulse, headache"),      # patient 6
  outcome = c("Recover", "Death", "Death", "Recover", "Recover", "Recover"))

datatable(df)
```


## Standardizing Strings

There are a lot of useful functions for manipulating strings we won't have time to demonstrate. You can find information on the use of these at [this link](https://epirhandbook.com/en/characters-and-strings.html#clean-and-standardise).


# Column cleaning

## Data

Let's load in a simulated ebola outbreak dataset to take a look at some other cleaning operations in R. 

```{r}
library(readxl)
linelist_raw <- read_excel("../data/linelist_raw.xlsx")
datatable(head(linelist_raw, 50))
```




