---
title: "Modeling and Testing"
format: 
  html: default
code-annotations: select
---

# Data

We import a dataset of cases from a simulated Ebola epidemic. 

```{r}
library(tidyverse)
library(DT)

```

# Distributions in R

Every probability distribution in R has four functions, with a root distribution name and a single letter prefix:

- p for "probability", the cumulative distribution function (c. d. f.)
- q for "quantile", the inverse c. d. f.
- d for "density", the density function (p. f. or p. d. f.)
- r for "random", a random variable having the specified distribution 


# Performing simple tests


# Regression

## Example Analysis - Infant Mortality 


## Example Analysis - Spline Regression with NHANES

### Load the data

There are 6063 observations, some are incomplete and have missing values for some covariates. In the metadata we have a textual description of the phenotype, the short name, and the target.  The target tells us which of the sampled individuals was eligible to answer the question. 

```{r loadData}
load("../data/nhanes_spline/d4.rda")
load("../data/nhanes_spline/metaD.rda")
datatable(metaD)
```



# High-performance modeling

These datasets were relatively small. If we want to analyze very large datasets, we might want to profile our analysis, checking it performance bottlenecks and memory usage.

```{r, eval = FALSE}
library(profvis)
profvis({
  
  # loess curve
  l2 = loess(lm1$residuals ~ lm1$fitted.values)
  pl = predict(l2, newdata=sort(lm1$fitted.values))

  # Spline regression
  lm2 = lm(d4$LBXTC ~ ns(d4$RIDAGEYR, df=7))
  summary(lm2)
  
  # Comparison
  anova(lm1, lm2)
  
  # PCA and processing
  cvars = c("RIDAGEYR", "INDFMPIR", "LBDHDD", "LBXGH", "BMXBMI", "LBXTC")
  cont_d = d4[, cvars]
  
  complete_ind = complete.cases(cont_d)
  
  cont_d = cont_d[complete_ind,]
  d4_missing = d4[!complete_ind,]
  d4_comp = d4[complete_ind,]
  
  pcs = prcomp(cont_d)
  
  which(abs(pcs$x[,1]) > max(abs(pcs$x[,1]))*0.95)
  which(abs(pcs$x[,2]) > max(abs(pcs$x[,2]))*0.95)
  
  # Random forest
  rf1 = randomForest(LBXTC ~ ., proximity=TRUE, data=cont_d)
})
```

We can take a look at some packages in R for high performance computing, such as packages for [data sets too large for memory](https://cran.r-project.org/web/views/HighPerformanceComputing.html)

# Analyzing Survey Data

```{r}
library(survey)
```

## Data

Let's take another look at the NHANES diabetes data we were using yesterday. 
This time we will perform an analysis using only a subset of the total dataset, focusing on adolescents. 

```{r}
all_nhanes <- read_csv("../data/nhanes_diabetes.csv")
nhanes_metadata <- read_csv("../data/nhanes_diabetes_metadata.csv")
datatable(nhanes_metadata)
```

## Preparing Survey Data

### Removing NA weights

We can't have any missing values in the survey design variables. 

Remove all rows with `NA` for the main survey design variables; 
`WTMEC2YR`, `SDMVPSU`, and `SDMVSTRA`.

```{r}
wt_nhanes <- all_nhanes %>%
  drop_na(WTMEC2YR, SDMVPSU, SDMVSTRA)
```


## Creating combined survey weights

Currently, our survey weights `WTSAF2YR` are for each 2 year cycle. 
We need to combine them to represent the full 6-year period we are investigating. 
Luckily NHANES has an [official guide](https://wwwn.cdc.gov/nchs/nhanes/tutorials/weighting.aspx) for combining these weights.
It turns out, all we need to do is divide all weights by 3. 

```{r}
wt_nhanes <- wt_nhanes %>%
  mutate(WTMEC6YR = WTMEC2YR * 1/3)
wt_nhanes=data.frame(wt_nhanes)
 wt_nhanes$diabetes = factor(wt_nhanes$diabetes, levels=c("nondiabetic","prediabetic","diabetic"))
```

## Creating the survey design object
We need to use some specialized survey analysis methods which are contained in the `survey` package written by Thomas Lumley.

```{r}
nhanes_design <- svydesign(id     = ~SDMVPSU,
                          strata  = ~SDMVSTRA,
                          weights = ~WTMEC6YR,
                          nest    = TRUE,
                          survey.lonely.psu = "adjust",
                          data    = wt_nhanes)

summary(nhanes_design)
```

Now we can take our data subset from the survey design object. 

```{r}
ado_design <- subset(nhanes_design, RIDAGEYR >= 13 & RIDAGEYR <= 18 & !is.na(OHXDECAY))

#Also make a tibble of this data to analyze
ado_data <- wt_nhanes %>%
  filter(RIDAGEYR >= 13 & RIDAGEYR <= 18) %>% # Gets the 3660 nonedentulous adolescents
  filter(!is.na(OHXDECAY)) %>% # Gets the 3346 with non-NA dental carie variable
  filter(!is.na(diabetes)) # Gets the 3046 with a diabetic status
```


# Exploring packages

## Predictive Modeling with TidyModels

Let's look at a quick predictive modeling example using Tidymodels. 

We can split the data to create training and testing sets. 

```{r}
library(tidymodels)

iris_split <- initial_split(iris, prop = 0.6)
iris_split
```

```{r}
iris_split %>%
  training() %>%
  glimpse()
```

We then create a **recipe**, a set of preprocessing steps we want to make for the data. 

```{r}
iris_recipe <- training(iris_split) %>%
  recipe(Species ~.) %>%
  step_corr(all_predictors()) %>%
  step_center(all_predictors(), -all_outcomes()) %>%
  step_scale(all_predictors(), -all_outcomes()) %>%
  prep()

iris_recipe
```

We then *bake* the recipe on the testing set, which runs the analysis steps on for the subset of the data. 

```{r}
iris_testing <- iris_recipe %>%
  bake(testing(iris_split)) 

glimpse(iris_testing)
```

Performing the same operation over the training data is redundant, because that data has already been prepped. To load the prepared training data into a variable, we use `juice()`. It will extract the data from the `iris_recipe` object.

```{r}
iris_training <- juice(iris_recipe)

glimpse(iris_training)
```

We can then train a model on the pre-processed data.

```{r}
iris_rf <-  rand_forest(trees = 100, mode = "classification") %>%
  set_engine("randomForest") %>%
  fit(Species ~ ., data = iris_training)
```

And finally create predictions on our test set and examine performance metrics. It is very easy to add the predictions to the baked testing data by using dplyr’s `bind_cols()` function.

```{r}
iris_rf %>%
  predict(iris_testing) %>%
  bind_cols(iris_testing) %>%
  glimpse()
```

```{r}
iris_rf %>%
  predict(iris_testing) %>%
  bind_cols(iris_testing) %>%
  metrics(truth = Species, estimate = .pred_class)
```

The above example comes from [here](https://rviews.rstudio.com/2019/06/19/a-gentle-intro-to-tidymodels/)

### Looking up vignettes

## Regularized regression with glmnet

```{r, eval=FALSE}
browseVignettes("glmnet")
```

## Bayesian Statistics with Stan and brms

```{r, eval=FALSE}
browseVignettes("brms")
```

# Other packages of note

- Deep learning with [`keras`](https://cran.r-project.org/web/packages/keras/vignettes/)
- [caret](https://topepo.github.io/caret/) for predictive modeling
- CRAN Task View on [machine learning](https://cran.r-project.org/web/views/MachineLearning.html)
- CRAN task view on [Bayesian Inference](https://cran.r-project.org/web/views/Bayesian.html)