---
title: "Summarizing and Visualizing data"
format: 
  html: default
#runtime: shiny
---

# Dataframes

We will begin with a mostly processed dataset from [NHANES](https://www.cdc.gov/nchs/nhanes/index.htm) designed to examine the relationship between diabetes and dental caries. Specifically, this dataset contains the 31034 records in NHANES from 2005 to 2010 with non-missing dental decay data.

Reference for dataframe inspection:

**Size**:

- `dim(nhanes)` - returns a vector with the number of rows as the first
  element, and the number of columns as the second element (the
  **dim**ensions of the object).
- `nrow(nhanes)` - returns the number of rows.
- `ncol(nhanes)` - returns the number of columns.

**Content**:

- `head(nhanes)` - shows the first 6 rows.
- `tail(nhanes)` - shows the last 6 rows.

**Names**:

- `names(nhanes)` - returns the column names (synonym of `colnames()` for
  `data.frame` objects).
- `rownames(nhanes)` - returns the row names.

**Summary**:

- `str(nhanes)` - structure of the object and information about the
  class, length and content of each column.
- `summary(nhanes)` - summary statistics for each column.

Note: most of these functions are "generic", they can be used on other types of objects besides `data.frame`.

# Tidyverse

```{r}
library(tidyverse)
library(DT)
```

You can find a [data transformation with dyplr cheatsheet here](https://rstudio.github.io/cheatsheets/data-transformation.pdf).

Instead of `read.csv()`, we will read in our data using the `read_csv()`
function (notice the `_` instead of the `.`), from the tidyverse package
**`readr`**.

Let's also load in the metadata for this dataset. 

## Exercise:

Using pipes, subset the `nhanes` data to keep female participants 15 years or older,
where the LBXGH is greater than 5.2 (and is not `NA`), and retain only the columns `RIAGENDR`, `RIDAGEYR`, and `LBXGLU`.

# Summarizing Data in Tables

# Visualizing Data

## First steps with ggplot2 

**ggplot2** is the most popular data visualisation R package. Its `ggplot()` function is at the core of this package, and this whole approach is colloquially known as *"ggplot"* with the resulting figures sometimes affectionately called "ggplots". The "gg" in these names reflects the "**g**rammar of **g**raphics" used to construct the figures. **ggplot2** benefits from a wide variety of supplementary R packages that further enhance its functionality.

The [data visualization with ggplot cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) from the RStudio website is a great reference to have on-hand when creating pltos. If you want inspiration for ways to creatively visualise your data, we suggest reviewing websites like the [R graph gallery](https://www.r-graph-gallery.com/) and [Data-to-viz](https://www.data-to-viz.com/caveats.html).

We can summarize the ggplot approach as:

```r
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()
```

## Interactive plots and widgets

We can use plotly to add basic interactivity to plots. 

```{r, eval=FALSE}
library(plotly)

p <- ggplot(nhanes, aes(x = BMXBMI, fill = RIAGENDR)) +
  geom_histogram(position = "stack") 

ggplotly(p)
```

And we can create more complex interactive plots using the **shiny** package. 

```{r, eval=FALSE}
library(shiny)
selectInput(
  'breaks', label = 'Number of bins:',
  choices = c(10, 20, 35, 50), selected = 20
)

renderPlot({
  ggplot(nhanes, aes(x = BMXBMI, fill = RIAGENDR)) +
  geom_histogram(position = "stack", bins = as.numeric(input$breaks)) 
  
})
```
This is just scratching the surface of interactive plots. You can find examples of creating full interactive dashboards in [The Epidemiologist R Handbook](https://epirhandbook.com/en/dashboards-with-r-markdown.html).

## Creating maps and using shape files

As with everything else, there are a variety of ways to create maps in R. Let's look at 2 examples, creating simple maps with the `maps` package and importing a shapefile using the `sf` package. 

For simple information we want to see on a standard map, we can use the `maps` package. 
For instance, we can grab a map of the USA and plot some coordinates.

```{r}
library(maps)
library(ggmap)
library(tmap)

usa <- map_data("usa") 

labs <- data.frame(
  long = c(-122.064873, -122.306417),
  lat = c(36.951968, 47.644855),
  names = c("SWFSC-FED", "NWFSC"),
  stringsAsFactors = FALSE
  )  

ggplot() +
  geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = "coral", color = "brown") + 
  coord_fixed(1.3) +
  geom_point(data = labs, aes(x = long, y = lat), color = "black", size = 5) +
  geom_point(data = labs, aes(x = long, y = lat), color = "gold", size = 4)
```
Or we can get data for states

```{r}
states <- map_data("state")
head(states)
```

And then grab and plot just new england. 

```{r}
new_england <- subset(states, region %in% c('connecticut', 'maine', 'massachusetts', 'new hampshire', 'rhode island', 'vermont'))

ggplot(data = new_england) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill = "palegreen", color = "black") + 
  coord_fixed(1.3)
```
Let's zoom in and plot population density by county in CT. 

```{r}
ct_df <- subset(states, region == "connecticut")

counties <- map_data("county")
ct_county <- subset(counties, region == "connecticut")
```

We start by plotting just the state boundaries. Let's get rid of the background gridlines too. 

```{r}
ct_base <- ggplot(data = ct_df, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "black", fill = "gray")
ct_base + theme_nothing()
```

Now we can add the county borders in white. 

```{r}
ct_base + theme_nothing() + 
  geom_polygon(data = ct_county, color = "white", aes(fill = subregion))
```
```{r}
library(sf)

linelist <- readRDS("../data/linelist_cleaned.rds")

# generate 1000 random row numbers, from the number of rows in linelist
sample_rows <- sample(nrow(linelist), 1000)

# subset linelist to keep only the sample rows, and all columns
linelist <- linelist[sample_rows,]
```

We use the package `sf` (spatial features) and its function `st_as_sf()` to create the new object we call `linelist_sf`. This new object looks essentially the same as the `linelist`, but the columns `lon` and `lat` have been designated as coordinate columns, and a coordinate reference system (CRS) has been assigned for when the points are displayed. 

```{r}
linelist_sf <- linelist |>
     sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)

datatable(head(linelist_sf, 10), rownames = FALSE, options = list(pageLength = 5, scrollX=T), class = 'white-space: nowrap' )
```

```{r}
sle_adm3_raw <- read_sf("../data/gis/sle_adm3.shp")

# ADM3 level clean
sle_adm3 <- sle_adm3_raw |>
  janitor::clean_names() |> # standardize column names
  filter(admin2name %in% c("Western Area Urban", "Western Area Rural")) # filter to keep certain areas
```

The package `tmap` offers simple mapping capabilities for both static (“plot” mode) and interactive (“view” mode) with just a few lines of code. 
We can plot the distribution of ebola cases using `tmap`.

```{r}
tmap_mode("plot") # choose either "view" or "plot"
tm_shape(linelist_sf) + tm_dots(size=0.08, col='blue')
```
This is not very useful without the administrative border data. 

```{r}
tmap_mode("plot") # choose either "view" or "plot"
tm_shape(sle_adm3,               # admin boundaries shapefile
           bbox = c(-13.3, 8.43,  # corner
                  -13.2, 8.51)) +   # corner
  tm_polygons(col = "#F7F7F7")+    # show polygons in light grey
  tm_borders(col = "#000000",      # show borders with color and line weight
             lwd = 2) +
  tm_text("admin3name")   +         # column text to display for each polygon
  tm_shape(linelist_sf) +
  tm_dots(size=0.08, col='blue', alpha = 0.5) +
  tm_layout(title = "Distribution of Ebola cases")   # give title to map
```

John A. Graves has a great example analysis in R using the Dartmouth atlas data to examine different definitions of health care markets which you can find [here](https://github.com/graveja0/health-care-markets).  

## Exporting plots

Exporting ggplots is made easy with the `ggsave()` function from **ggplot2**. It can work in two ways, either:

-   Specify the name of the plot object, then the file path and name with extension
    -   For example: `ggsave(my_plot, "plots/my_plot.png"))`\
-   Run the command with only a file path, to save the last plot that was printed
    -   For example: `ggsave("plots/my_plot.png"))`

You can export as png, pdf, jpeg, tiff, bmp, svg, or several other file types, by specifying the file extension in the file path.

You can also specify the arguments `width =`, `height =`, and `units =` (either "in", "cm", or "mm"). You can also specify `dpi =` with a number for plot resolution (e.g. 300). See the function details by entering `?ggsave` or reading the [documentation online](https://ggplot2.tidyverse.org/reference/ggsave.html).

------------------------------------------------------------------------

*The materials in this lesson have been adapted from work created by the (HBC)\](http://bioinformatics.sph.harvard.edu/) and Data Carpentry (http://datacarpentry.org/), as well as materials created by Laurent Gatto, Charlotte Soneson, Jenny Drnevich, Robert Castelo, and Kevin Rue-Albert. These are open access materials distributed under the terms of the [Creative Commons Attribution license](https://creativecommons.org/licenses/by/4.0/) (CC BY 4.0), which permits unrestricted use, distribution, and reproduction in any medium, provided the original author and source are credited.*